Stage-Splitting a Modal Language

### 2014-09-28 paper/semantics.tex
Editing pass in Section 2 and Section 3.0.

\section {Staged Programming in \lang}
\label{sec:staging}
\begin{figure*}
\label{fig:quickselect}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) 
  : (int*list*list) =
  case unroll l of 
    Empty => (0,Empty, Empty) 
  | Cons (h,t) =>
      let (s,left,right) = partition (p,t) in
      if h<p 
      then (s+1,Cons(h,left),right)
      else (s,left,Cons(h,right))

fun qs (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qs (left, k)
        | EQ => h
        | GT => qs (right, k-n-1)
\end{lstlisting}
\caption{Unstaged Code}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qs (l : list, k : $int) : $int = 
  case l of
    1`Empty` => next {2`0`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        let n = hold{n} in
          case compare prev{k} n of
            LT => prev {qs (left k)}
          | EQ => hold {h}
          | GT => prev {qs (right, next{prev{k}-n-1)}}
      }	
\end{lstlisting}
\caption{Staged Code}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

@@

fun qs2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qs2 (p1,k)
      | EQ => h
      | GT => qs2 (p2,k-n-1)
\end{lstlisting}
\caption{Split Code}
\end{minipage}
\caption{Caption place holder}
\end{figure*}

For an example of an algorithm amenable to staging techniques,
consider the quickselect algorithm in [figure part (a)],
which finds the $k$th largest element of a list.
It operates by inspecting the head of the list, 
partitioning the rest of the list by the head,
comparing the size of the partition to the desired index,
and the recursively selecting on the correct sublist.
Assuming that the list is randomly sorted, 
quickselect will take expected $O(n)$ time, where $n$ is the size of the list.

Importantly, quickselect has property that the partitioning of the list does not depend on $k$, 
even through the recursive boundary.\footnote{There is a control-dependence on $k$, but we can ignore that for now.}
We can leverage this dependency pattern to split quickselect into two other functions:
one which precomputes all of the partitions of the input list to build a binary search tree,
and another which uses this tree to perform an accelerated index lookup.
This split version of quickselect is shown in [figure part (c)].
One advantage of this form is that the tree can be built once per list
and then used to accelerate many lookups.

The goal of our work is to automate this splitting transformation.
Our first step is to define a language which clearly and unambiguously specifies how its terms should be split.
Ideally, only minimal refactoring should be required to write terms within this language,
and from there, the type system should be enough to prove that a valid splitting exists.

To this end, we present \lang.
The main idea of \lang\ is that all parts of terms defined in it can be identified with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.
Correspondingly, after splitting those parts of the \lang\ term in \bbone\ will end up in the precomputation, 
and those parts in \bbtwo\ will end up in the residual.
Within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.  
Indeed this is a necessary property if we want to be able to split the term.

[Figure part (b)] shows a staged implementation of quickselect in \lang.
As before, the inputs to the function are the list to select from and the index to select,
but now the latter is represented with the type $\fut\rmint$ rather than $\rmint$.
The difference here is that an $\rmint$ is an integer available at the current stage (stage \bbone), 
whereas a $\fut\rmint$ is an integer available only at the next stage (stage \bbtwo).
Naturally the output type, representing the $k$th largest element of the list,
is also $\fut\rmint$, since it cannot be known until the next stage.

Each construct in the body of quickselect is now associated with a stage via an interleaving of $\next$ and $\prev$ blocks.  
Specifically, $\next$ occurs in a stage \bbone\ context and indicates that the contents of its block are stage \bbtwo, 
whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
(We adapt the convention convention that the top-level context is stage \bbone.)
The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  
For example, \verb|next{0}| from above has the type $\fut\rmint$.
Correspondingly, $\prev$ requires that its stage \bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
For example, \verb|prev{k}| from above has type $\rmint$ at stage \bbtwo, since $k$ is bound to a $\fut\rmint$ at stage \bbone.
These type restrictions essentially enforce that ``later stage content" is always treated hygienically at stage \bbone,
which is necessary to admit a properly staged implementation.

The code also contains two $\pause$ blocks.  
This construct has same stage signature as $\prev$,
but instead of ``unwrapping" $\fut$ types it simply promotes integers from stage \bbone\ to stage \bbtwo.
\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
We instead provide it as a primitive to shorten examples.  
Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
This is related to the notion of {\em mobility} in \cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}

But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
the staged version of quickselect is virtually identical to the unstaged version.
The constructs that were added were placed in order to maximize the work done at stage \bbone\ while still conforming to the type signature.
It would have also been valid to simply move the whole input list unchanged into stage \bbtwo\ at the very beginning, 
but that would not be particularly interesting since it would result in an effectively trivial split 
that's just the identity at stage \bbone\ and plus quickselect at stage \bbtwo.
There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
This process is known as {\em binding time analysis}, and we do not consider it here.
Instead, we assume that all input programs are properly staged according to some programmer's intent.

@@

As stated in \cref{sec:staging}, we operate on a language called \lang,
which is a typed two-stage lambda calculus featuring products, sums, and isorecursive types.
The grammar of \lang\ can be found in \cref{fig:grammar}.

\subsection{Statics}
The grammar and type system of \lang\ are a simple adaptation of those of \cite{davies96}, 
restricted to two stages\footnote{This restriction to two stages is made primarily for 
simplicity of presentation.  All of these techniques could feasibly be extended to more
complicated stage systems.} and extended with general sums and recursion.

As explained before, the $\fut$ constructor is a way to encapsulate a stage \bbtwo\ type at stage \bbone.  
But because of the two-stage assumption, $\fut$ has no meaning when used at stage \bbtwo.
This restriction is managed by the type validity judgment, whose rules are given in \cref{fig:validTypes}. 
In short, $\Delta \vdash A \istypewor$ means that type $A$ is valid at world $w$ under the assumptions in $\Delta$.
This context is only augmented by recursive types.

@@

A key attribute of the quickselect example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. 
Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside stage \bbtwo\ ones 
(such as the recurive call to {\tt fexp} above), 
this strategy conflicts with the precept of staged execution: 
that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.

Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
subexpressions before any of its stage \bbtwo\ subexpressions. This results in a
stage \bbtwo\ term with no stage \bbone\ content; we say this is a term in a
monostaged language called \langTwo. Then, we perform the remainder of the evaluation with 
$\tworedsym$, an ordinary dynamic semantics for \langTwo\ (the rules for this
judgment are not shown, but they are standard).

>>>>

\section {Staged Programming in \lang}
\label{sec:staging}

\begin{figure*}
\label{fig:quickselect}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) 
  : (int*list*list) =
  case unroll l of 
    Empty => (0,Empty, Empty) 
  | Cons (h,t) =>
      let (s,left,right) = partition (p,t) in
      if h<p 
      then (s+1,Cons(h,left),right)
      else (s,left,Cons(h,right))

fun qs (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qs (left, k)
        | EQ => h
        | GT => qs (right, k-n-1)
\end{lstlisting}
\caption{Unstaged implementation of quickselect.}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qs (l : list, k : $int) : $int = 
  case l of
    1`Empty` => next {2`0`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        let n = hold{n} in
          case compare prev{k} n of
            LT => prev {qs (left k)}
          | EQ => hold {h}
          | GT => prev {qs (right, next{prev{k}-n-1)}}
      }	
\end{lstlisting}
\caption{Staged implementation of quickselect in \lang.}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{lstlisting}
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

@@

fun qs2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qs2 (p1,k)
      | EQ => h
      | GT => qs2 (p2,k-n-1)
\end{lstlisting}
\caption{Split (two-phase) implementation of quickselect.}
\end{minipage}
\caption{Caption place holder}
\end{figure*}

As an example of an algorithm amenable to staging techniques,
consider the quickselect algorithm shown in Figure~2a.
Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
then recursing on the sublist containing the desired element.

Importantly, one property of quickselect is that the partitioning of the list, even across recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  This independence allows the quickselect algorithm to be split into two functions:
one which precomputes all partitions of the input list (to construct a binary search tree---see \texttt{qs1} in Figure~2c),
and another which uses this tree to perform an accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qs2}).
One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
and then used to accelerate many future lookups for any value of $k$.

This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.  \lang clearly and unambiguously specifies how its terms should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each term with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.  After splitting, those parts of the \lang\ term in stage \bbone\ will end up in the precomputation, 
and those parts in stage \bbtwo\ will end up in the residual.
Within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.  
Indeed this is a necessary property if we want to be able to split the term.

Figure 2b shows a staged implementation of quickselect in \lang.
As before, the function accepts as input the list to select from and a value of $k$,
but now the latter is represented with the type $\fut\rmint$ rather than $\rmint$.
The difference here is that an $\rmint$ is an integer available at the current stage (stage \bbone), 
whereas a $\fut\rmint$ is an integer available only at the next stage (stage \bbtwo).
Naturally the output type, representing the $k$th largest element of the list,
is also $\fut\rmint$, since it is not computed until stage \bbtwo.

Each construct in the body of quickselect is now associated with a stage via an interleaving of $\next$ and $\prev$ blocks.  
Specifically, $\next$ occurs in a stage \bbone\ context and indicates that the contents of its block are stage \bbtwo, 
whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
(We adapt the convention that the top-level context is stage \bbone.)
The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  
For example, \verb|next{0}| from above has the type $\fut\rmint$.
Correspondingly, $\prev$ requires that its stage \bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
For example, \verb|prev{k}| from above has type $\rmint$ at stage \bbtwo, since $k$ is bound to a $\fut\rmint$ at stage \bbone.
These type restrictions essentially enforce that ``later stage content" is always treated hygienically at stage \bbone,
which is necessary to admit a properly staged implementation.

The code also contains two $\pause$ blocks.  
This construct has same stage signature as $\prev$,
but instead of ``unwrapping" $\fut$ types it simply promotes integers from stage~\bbone\ to stage~\bbtwo.
\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
We instead provide it as a primitive to shorten examples.  
Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
This is related to the notion of {\em mobility} in \cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}

But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
the staged version of quickselect in Figure~2b is virtually identical to the unstaged version in Figure~2a.
The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.
It would have also been valid to simply move the whole input list unchanged into stage~\bbtwo\ at the very beginning, 
but that would not be particularly interesting since it would result in an effectively trivial split 
that's just the identity at stage~\bbone\ and plus quickselect at stage~\bbtwo.
There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
This process is known as {\em binding time analysis}, and we do not consider it here.
Instead, we assume that all input programs are properly staged according to some programmer's intent.

@@

As stated in \cref{sec:staging}, in this paper we describe splitting in the context of \lang,
a typed two-stage lambda calculus featuring products, sums, and isorecursive types.
The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of \cite{davies96}, 
restricted to two stages and extended with general sums and recursion\footnote{This restriction to two stages is made primarily for 
simplicity of presentation.  All of these techniques could feasibly be extended to more
complicated stage systems.}.  The following subsections provide a brief overview of the type system for \lang, and highlight important differences in its semantics from this prior work.

\subsection{Statics}

In \lang, the $\fut$ constructor encapsulates a stage \bbtwo\ type at stage~\bbone.  
But because of the two-stage assumption, $\fut$ has no meaning when used at stage~\bbtwo.
This restriction is managed by the type validity judgment, whose rules are given in \cref{fig:validTypes}. 
In short, $\Delta \vdash A \istypewor$ means that type $A$ is valid at world $w$ under the assumptions in $\Delta$.
This context is only augmented by recursive types.

@@

A key attribute of the quickselect example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. 
Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside stage~\bbtwo\ ones 
(such as the recurive call to {\tt fexp} above), 
this strategy conflicts with the precept of staged execution: 
that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.

Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
subexpressions before any of its stage~\bbtwo\ subexpressions. This results in a
stage \bbtwo\ term with no stage~\bbone\ content; we say this is a term in a
monostaged language called \langTwo. Then, we perform the remainder of the evaluation with 
$\tworedsym$, an ordinary dynamic semantics for \langTwo\ (the rules for this
judgment are not shown, but they are standard).
